Forecasting biodiversity in breeding birds using best practices

### 2018-01-05 manuscript/draft.Rmd
Editor comments: pp 1-9

Forecasting the future state of ecological systems is increasingly important for
planning and management, and also for quantitatively evaluating how well
ecological models capture the key processes governing natural systems
[@clark2001; @houlahan2017; @dietze2017]. Forecasts regarding biodiversity are
especially important, due to biodiversity's central role in conservation
planning and its sensitivity to anthropogenic effects [@cardinale2012;
@diaz2015; @tilman2017]. High-profile studies forecasting large biodiversity
declines over the coming decades have played a large role in shaping ecologists'
priorities [as well as those of policymakers; e.g. @ipcc2014], but it is
inherently difficult to evaluate such long-term predictions before the projected
biodiversity declines have occurred.

Previous efforts to predict future patterns of terrestrial species richness, and
diversity more generally, have focused primarily on building species
distributions models [SDMs; @thomas2004; @thuiller2011; @urban2015]. In general,
these models describe individual species' occurrence patterns as functions of
the environment. Given forecasts for environmental conditions, these models can
predict where each species will occur in the future. These species-level 
predictions are then combined ("stacked") to generate forecasts for species 
richness [e.g. @calabrese2014]. Alternatively, models that directly relate
spatial patterns of species richness to environment conditions have been
developed and generally perform equivalently to stacked SDMs [@algar2009;
@distler2015]. This approach is sometimes referred to as "macroecological"
modeling, because it models the larger-scale pattern (richness) directly
[@distler2015].

@@

**Richness data.** Bird species richness was obtained from the North American
Breeding Bird Survey (BBS) [@pardieck2017] using the Data Retriever Python
package [@morris2013; @senyondo2017] and rdataretriever R package
[@rdataretriever]. BBS observations are three-minute point counts made at 50
fixed locations along a 40km route. Here we denote each route as a site and
summarize richness as the total species observed at all 50 locations in each
surveyed year. Prior to summarizing the data was filtered to exclude all
nocturnal, cepuscular, and aquatic species [since these species are not well
sampled by BBS methods; @hurlbert2005], as well as unidentified species, and
hybrids. All data from surveys that did not meet BBS quality criteria were also
excluded.

@@

**Future environmental projections.** In addition to the analyses presented
here, we have also generated and archived long term forecasts from 2014-2050.
This will allow future researchers to assess the performance of our six models
on longer time horizons as more years of BBS data become available.
Precipitation and temperature were forecast using the CMIP5 multi-model ensemble
dataset [@brekke2013]. 37 downscaled model runs [@brekke2013, see Table S1]
using the RCP6.0 scenario were averaged together to create a single ensemble
used to calculate the bioclimatic variables for North America. For NDVI, we used
the per-site average values from 2000-2013 as a simple forecast. For observer
effects (see below), each site was set to have zero observer bias. The
predictions have been archived at [@zenodoPredict].

@@

This research was supported by the Gordon and Betty Moore Foundation's
Data-Driven Discovery Initiative through Grant GBMF4563 to E.P. White. We thank
the developers and providers of the data and software that made this research
possible including: the PRISM Climate Group at Oregon State University, the
staff at USGS and volunteer citizen scientists associated with the North
American Breeding Bird Survey, NASA, the World Climate Research Programme's
Working Group on Coupled Modelling and its working groups, the U.S. Department
of Energy's Program for Climate Model Diagnosis and Intercomparison, and the
Global Organization for Earth System Science Portals. A. C. Perry provided
valuable comments that improved the clarity of this manuscript.

>>>>

Forecasting the future state of ecological systems is increasingly important for
planning and management, and also for quantitatively evaluating how well
ecological models capture the key processes governing natural systems
[@clark2001; @houlahan2017; @dietze2017]. Forecasts regarding biodiversity are
especially important, due to biodiversity's central role in conservation
planning and its sensitivity to anthropogenic effects [@cardinale2012;
@diaz2015; @tilman2017]. High-profile studies forecasting large biodiversity
declines over the coming decades have played a large role in shaping the
priorities of ecologists and policymakers [e.g. @ipcc2014], but it is inherently
difficult to evaluate such long-term predictions prior to the predicted decline
in diversity.

Previous efforts to predict future patterns of terrestrial species richness, and
diversity more generally, have focused primarily on building species
distributions models [SDMs; @thomas2004; @thuiller2011; @urban2015]. In general,
these models describe individual species' occurrence patterns as functions of
the environment. Given forecasts for environmental conditions, these models can
predict where each species will occur in the future. These species-level
predictions are then combined ("stacked") to generate forecasts for species
richness [e.g. @calabrese2014]. Alternatively, models that directly relate
spatial patterns of species richness to environment conditions are available
developed and generally perform equivalently to stacked SDMs [@algar2009;
@distler2015]. This approach is sometimes referred to as "macroecological"
modeling, because it models the larger-scale pattern (richness) directly
[@distler2015].

@@

**Richness data.** Bird species richness was obtained from the North American
Breeding Bird Survey (BBS) [@pardieck2017] using the Data Retriever Python
package [@morris2013; @senyondo2017] and rdataretriever R package
[@rdataretriever]. BBS observations are three-minute point counts made at 50
fixed locations along a 40km route. Here we denote each route as a site and
summarize richness as the total species observed at all 50 locations in each
surveyed year. Prior to summarizing the data was filtered to exclude all
nocturnal, crepuscular, and aquatic species [since these species are not well
sampled by BBS methods; @hurlbert2005], as well as unidentified species, and
hybrids. All data from surveys that did not meet BBS quality criteria were also
excluded.

@@

**Future environmental projections.** In addition to the analyses presented
here, we also generated and archived long term forecasts from 2014-2050. This
will allow future researchers to assess the performance of our six models on
longer time horizons as more years of BBS data become available. Precipitation
and temperature were forecast using the CMIP5 multi-model ensemble dataset
[@brekke2013]. We averaged 37 downscaled model runs [@brekke2013, see Table S1]
using the RCP6.0 scenario to create a single ensemble used to calculate the
bioclimatic variables for North America. For NDVI, we used the per-site average
values from 2000-2013 as a simple forecast. For observer effects (see below),
each site was set to zero observer bias. The predictions were archived at
[@zenodoPredict].

@@

We thank the developers and providers of the data and software that made this
research possible including: the PRISM Climate Group at Oregon State University,
the staff at USGS and volunteer citizen scientists associated with the North
American Breeding Bird Survey, NASA, the World Climate Research Programme's
Working Group on Coupled Modelling and its working groups, the U.S. Department
of Energy's Program for Climate Model Diagnosis and Intercomparison, and the
Global Organization for Earth System Science Portals. A. C. Perry provided
valuable comments that improved the clarity of this manuscript.
